---
title: "Project LSTAT 2130"
author:
- De Rongé Jessica
- Termont Didier
- Visschers Antoine
output:
  pdf_document: 
    toc: true
    toc_depth: 2
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,comment = NA, tidy="styler")
```

```{r libraries, include=FALSE}
library(ggplot2)
library(coda)
library(rjags)
library(R2WinBUGS)

```


```{r data}
cannabis <- read.csv("cannabis.txt",sep="",header = TRUE)
set.seed(20220501)
```

```{r functions_def}
M <- 10000

log.f <- function(pi,N,xplus) {
  gamma <- (1+35*pi)/36
  (xplus*log(gamma)) + ((N-xplus)*log(1-gamma))
}

log.f.2 <- function(pi,N,xplus) {
  (N-xplus)*log(1-pi)+xplus*log(1+35*pi)
}

metropolis <- function(pi0,M, sd.prop, lpost, N, xplus) {
  pi = numeric(M)
  pi[1]=pi0
  
  n.accept = 0
  for(i in 2:M) {
    pi.prop = pi[i-1]+rnorm(1,0,sd.prop)
    prop = min(1,exp(lpost(pi.prop,N,xplus) - lpost(pi[i-1],N,xplus)))
    accept = (runif(1) <= prop)
    if(accept) {
      n.accept = n.accept+1
      pi[i] = pi.prop
    } else {
      pi[i] = pi[i-1]
    }
  }
  
  accept.rate = round(n.accept/(M-1),2)
  cat("Acceptance rate: ",accept.rate,"\n")
  
  return(pi)
}

```

\newpage

# Question 1

we have X the random variable for the answer of the question __Have you smoked cannabis?__ 

D is the event __rolled a double 6__. From the total sum rules we can write
$$P(X)=P(X|D)P(D)+P(X|\bar{D})P(\bar{D})$$

From the information we have we know that:

- $P(D)=\frac{1}{36}$
- $P(\bar{D})=\frac{35}{36}$
- $P(X=1|D) = 1$ and $P(X=0|D)=0$
- $Y|\bar{D} \sim Ber(\pi)$

Putting all together we have
$$P(X=x) = \frac{x}{36} + \frac{35}{36} (1-\pi)^{1-x} \pi^x$$

We want to compute the probability $\gamma$ that the awnser is __yes__
$$\gamma = P(X=1) = \frac{1}{36} + \frac{35}{36} (1-\pi)^{1-1} \pi^1$$
$$\gamma = \frac{1+35\pi}{36}$$

We can state that X is folowing a Bernoulli distribution with parameter $\gamma$: $$X|\pi \sim Ber(\gamma) = Ber(\frac{1+35\pi}{36})$$

\newpage

# Question 2

We have $Y$ a new random variables that is the sum of all __yes__ anwser for a sample of size $n$ of iid variables:
$$Y = \sum_{i=1}^{n}{X_i}$$
with $X_i \sim Ber(\frac{1+35\pi}{36})$

from Bayes rules, we can write $$P(\pi|Y) \propto P(Y|\pi) P(\pi)$$

## Prior $P(\pi)$
we assume that we have no prior information on the proportion of persons that have smoken cannabis. Therefore we take $\pi \sim Uni(0,1)$.

$P(\pi) \propto 1$

## Likelihood $P(Y|\pi)$
Since all $X_i$ are iid we can write
$$P(Y|\pi) = \prod_{i=1}^{n}{P(X_i|\pi)} \propto \prod_{i=1}^{n}{(1-\gamma(\pi))^{1-x_i}\gamma(\pi)^{x_i}}$$

$$P(Y|\pi)\propto (1-\gamma(\pi))^{n - \sum{x_i}} \gamma(\pi)^{\sum{x_i}}$$

## Posterior $P(\pi|Y)$

From Bayes we know that $P(\pi|Y) \propto likelihood \times  prior$.
Since we have computed the likelihood and the prior, we have the posterior:

$$P(\pi|Y) \propto (1-\gamma(\pi))^{n - \sum{x_i}} \gamma(\pi)^{\sum{x_i}}$$

after replacing the $\gamma$ by $\pi$, we have a posterior that is proportionnal to $$ P(\pi|Y) \propto (1+35\pi)^{\sum{x_i}} (1-\pi)^{n-\sum{x_i}}$$

\newpage

# Question 3

## Metropolis algorithm for a sample of pi

```{r metro_cannabis,echo=TRUE}
pi = metropolis(
  pi0=0.5,M,sd=0.01,
  lpost=log.f.2,N=nrow(cannabis),xplus=sum(cannabis$y))
```

The following figure will give the convergence information of the sample:

```{r}
plot(pi)
```


## Plausible value for pi

for this we take a 95% credible interval on the sample for pi. We must remove the burn-in part of the sample
```{r ic95_cannabis, echo=TRUE}
quantile(pi[200:M],c(0.025,0.975))
```

## Probability that the proportion of cannabis smoker is over 10%

```{r, echo=TRUE}
nrow(subset(data.frame(pi[200:M]),pi>=0.1))/(M-200)
```

\newpage

# Question 4

We generate two samples with the metropolis algorithm. One for the proportion of male smokers (`pi_m`) and the other for the female smokers (`pi_f`)
The sample for the difference of proportion between the male and the female will be given by $\delta = pi_m[200:M] - pi_f[200:M]$.
We do not take the burn-in size (200 first elements of the samples)

```{r, echo=TRUE}
can_m =subset(cannabis,male==1)
pi_m = metropolis(
  pi0=0.5,M,sd.prop=0.009,
  lpost=log.f.2,N=nrow(can_m),xplus=sum(can_m$y))

can_f=subset(cannabis,male==0)
pi_f = metropolis(
  pi0=0.5,M,sd.prop=0.01,
  lpost=log.f.2,N=nrow(can_f),xplus=sum(can_f$y))

delta = pi_m[200:M] - pi_f[200:M]
```

The two next figures shows that the samples are converging to random variables and this help to generate the distribution of the $\delta$.

```{r, out.width="50%"}
plot(pi_m,ylim=c(0,0.5))
points(pi_f,col="red")
hist(delta)
```
With these we can have a first impression that $\delta > 0$. We can formalize this by computing the 95% credible interval and the probality $P(\delta > 0)$

IC:
```{r, echo=T}
quantile(delta,c(0.025,0.975))
```

$P(\delta>0)$
```{r, echo=T, comment=NA}
nrow(subset(data.frame(delta),delta>0))/length(delta)
```

We can conclude that there is a significant difference of cannabis smoker based on the gender. The proportion of male smokers is larger than the female smokers.

\newpage

# Question 5

## -a- MCMC in jags

We will use priors that are uninformative: $\alpha_0 , \alpha_1, \beta_0, \beta_1 \sim N(0,\sigma^2 = 10^6)$

The modeling of the logistic regression will be done by a Bernoulli experiment. 
From the question 1, we know that the experiment Y follows a Bernoulli with the parameter $\gamma = \frac{1+35\pi}{36}$
So we will have:

- $Y_i|\alpha_0,\alpha_1,\beta_0,\beta_1 \sim Ber(\frac{1+35\pi_i}{36})$
- $logit(\pi_i) = (\alpha_0 + \alpha_1(x-40)) \times male_i + (\beta_0 + \beta_1(x-40)) \times (1-male_i)$
- $x_i$ is the age of the person
- $male_i = 1$ if the person is a male, 0 otherwise

All of this is implemented in the following R code that will be used with 'rjags'

```{r mcmc_model, echo=TRUE, warning=FALSE}
model <- function() {
  #sample
  for(i in 1:n) {
    y[i] ~ dbin((1+35*pi[i])/36,1)
    logit(pi[i]) <- (
      (alpha0 + alpha1*(age[i]-40))*male[i]) + 
      ((beta0 + beta1*(age[i]-40))*(1-male[i]))
  }
  
  # prior
  alpha0 ~ dnorm(0,1E-6)
  alpha1 ~ dnorm(0,1E-6)
  beta0  ~ dnorm(0,1E-6)
  beta1  ~ dnorm(0,1E-6)
  
  # function to monitor
  delta <- alpha1 - beta1
  
  # predict the odds for males who are 25 year old
  etha_m_25 <- alpha0 + alpha1*(25-40)
}


write.model(model,"cannabis.bug")

```

Let's notice the part with `delta` and `etha_m_25` are here to answer to the questions 5.b and 5.c and will be discussed at that time.

The execution to generate the prior sample is done with the following code:

```{r mcmc, echo=TRUE}
cannabis.list <- with(cannabis,
                      list(y=y,
                           age=age,
                           male=male,
                           n=nrow(cannabis)))

params <- c("alpha0","alpha1","beta0","beta1","delta","etha_m_25")

inits.list <- list(list(
  alpha0=0,
  alpha1=0,
  beta0=0,
  beta1 =0),
  list(
    alpha0=-4,
    alpha1=-1,
    beta0=-5,
    beta1=-1
  )
  )


cannabis.model <- jags.model(
  file="cannabis.bug",
  data=cannabis.list,
  inits = inits.list,
  n.chains = length(inits.list))

update(cannabis.model,1000)

out = coda.samples(model=cannabis.model,
                   variable.names = params,
                   n.iter=20000)

out.matrix = as.matrix(out)

# plot(out)
```
 
```{r}
effectiveSize(out)
summary(out)
HPDinterval(out)
```
## -b- Plausible values

### $\alpha_1$

From the MCMC chains we can generate the histogram for $\alpha_1$

```{r, out.width="75%"}
hist(out.matrix[,"alpha1"],xlab="alpha1")
```

And extract the HPD interval at 95%:

```{r, echo=T}
HPDinterval(mcmc(out.matrix[,"alpha1"]))
```

### $\beta_1$
From the MCMC chains we can generate the histogram for $\beta_1$

```{r, out.width="75%"}
#ggplot(data=data.frame(out.matrix[,"beta1"]),aes(x=beta1)) +
#  geom_histogram()

hist(out.matrix[,"beta1"],xlab="beta1")
```

And extract the HPD interval at 95%:

```{r, echo=T}
HPDinterval(mcmc(out.matrix[,"beta1"]))
```

### $\delta = \alpha1 - \beta1$

### evolution of the odds $\eta_x$


## -c- probability of canabis usage for male of 25 years


